---
title: "IAT: Sexuality"
author:
  - name: Seungju Kim
    orcid: 0009-0009-5166-9170
    email: seungju7@illinois.edu
    affiliation: 
      - name: University of Illinois at Urbana-Champaign
        department: Psychology
format: 
  html:
    embed-resources: true
execute:
  eval: false
  echo: true
---

```{r}
#| include: false
pacman::p_load(DT)
```


## Source Description

**Full name:** Implicit Association Test - Sexuality\
**Website:** [Harvard Project Implicit](https://implicit.harvard.edu/implicit/selectatest.html)\
**Data type:** Survey\
**Unit of analysis:** State\
**Geographic Coverage:** 50 U.S. states + D.C.\
**Temporal Coverage:** 2016–2025

### About the Original Data

Project Implicit (housed within Harvard University) is a web-based platform that collects implicit associations and explicit attitudes toward sexuality. Data collection for the sexuality IAT has been ongoing since 2004 using a non-probability sample of volunteer participants who visit the website through search engines, social media, or school/work referrals.

::: callout-warning
Researchers should be aware of one significant limitation of the implicit associations measure for sexuality: it only captures implicit associations between good/bad and **gay men** (via the D-score). This means the implicit measure may not hold meaningful implications for lesbian, bi/pansexual, or otherwise sexually diverse individuals.
:::

### Original Data Collection

| Feature | Details |
|------------------------------------|----------------------------------------|
| Collection method | Continuous online survey (self-selected) |
| Sampling frame | U.S. adult internet users |
| Sample type | Convenience (non-probability) |
| Update frequency | Annual public data releases |
| Geographic coverage | State-level (U.S. only) |
| Temporal coverage | 2004–present |

::: callout-note
Project Implicit releases public-use datasets annually. File formats vary by year — earlier years are released as `.sav` (SPSS), later years as `.csv`. Variable naming conventions also shift across years (e.g., `tgaywomen` → `tgayleswomen` starting in 2017).
:::

### Required Citation

If you use this indicator in published research, you must cite the original data source in addition to the stigmaR package:

> Xu, K., Nosek, B., & Greenwald, A. (2014). Psychology data from the Race Implicit Association Test on the Project Implicit Demo website. *Journal of Open Psychology Data, 2*(1). https://doi.org/10.5334/jopd.ac

------------------------------------------------------------------------

## Access and Download

| Feature | Details |
|------------------------------------|----------------------------------------|
| Access type | Public download |
| URL | [OSF — Project Implicit Public Data](https://osf.io/y9hiq/) |
| Authentication required | No |
| Cost | Free |
| Last accessed | January 2026 |

### Download Procedure

Annual sexuality IAT datasets are available on the Project Implicit OSF page. Files are organized by year and released in `.csv` or `.sav` format.

1. Navigate to the OSF repository linked above.
2. Download one file per year for the Sexuality IAT (2016–2025).
3. Place files in `data/raw_data/iat/sexuality/`, organized by year.

::: callout-warning
2015 data uses a different variable set — it is missing several policy items (`adoptchild`, `serverights`, `transgender`, `marriagerights_3num`, `relationslegal_3num`) that are required for the composite indices. **2015 is excluded from this pipeline.** Pre-2015 data also differ substantially in structure and are not used.
:::

------------------------------------------------------------------------

## Cleaning Procedure

### Raw Data Structure

- **File format:** `.csv` (2016+) and `.sav` (select years)\
- **Rows:** One row per respondent\
- **Key variables used:** `state`, `year`, `d_biep.straight_good_all`, `att_7`, `tgaymen`, `tgayleswomen` / `tgaywomen`, `marriagerights_3num`, `relationslegal_3num`, `adoptchild`, `serverights`, `transgender`

### Cleaning Steps

#### 1. File Loading and Year Exclusion

All `.csv` and `.sav` files in the raw data folder are read in and merged into a single named list. The 2015 file is excluded due to incompatible variable structure.

``` r
csv_files <- csv_files[!grepl("2015", csv_files)]
```

Column names are normalized to lowercase to handle inconsistent casing across years and file types.

``` r
names(df) <- tolower(names(df))
```

#### 2. Variable Construction and Scoring

Nine item-level variables are constructed, each scored so that **higher values = more stigma** (anti-gay/lesbian attitudes).

A helper function `grab()` is used throughout to safely retrieve columns that may not exist in all years — returning `NA` if the variable is absent rather than throwing an error.

``` r
grab <- function(df, col) {
  if (col %in% names(df)) df[[col]] else rep(NA_real_, nrow(df))
}
```

```{r}
#| eval: true
#| echo: false
#| column: page
library(DT)

items_dict <- data.frame(
  Variable = c(
    "iat_sex_imp_d",
    "iat_sex_exp_att",
    "iat_sex_exp_therm_gm",
    "iat_sex_exp_therm_gw",
    "iat_sex_exp_pol_marr",
    "iat_sex_exp_pol_legal",
    "iat_sex_exp_pol_adopt",
    "iat_sex_exp_pol_serv",
    "iat_sex_exp_pol_trans"
  ),
  Label = c(
    "Implicit D-score",
    "Explicit attitude toward gay people",
    "Feeling thermometer: gay men (reversed)",
    "Feeling thermometer: lesbian women (reversed)",
    "Policy: marriage equality",
    "Policy: legal recognition of same-sex relations",
    "Policy: same-sex adoption rights",
    "Policy: right to refuse service",
    "Policy: transgender bathroom access"
  ),
  Type = c(
    "Implicit",
    "Explicit — Attitude",
    "Explicit — Thermometer",
    "Explicit — Thermometer",
    "Explicit — Policy",
    "Explicit — Policy",
    "Explicit — Policy",
    "Explicit — Policy",
    "Explicit — Policy"
  ),
  Source_Variable = c(
    "d_biep.straight_good_all",
    "att_7",
    "tgaymen",
    "tgayleswomen / tgaywomen",
    "marriagerights_3num",
    "relationslegal_3num",
    "adoptchild",
    "serverights",
    "transgender"
  ),
  Scoring = c(
    "Higher = stronger pro-straight implicit bias",
    "Higher = more positive toward straight (more stigma)",
    "10 − raw score; higher = more stigma",
    "10 − raw score; higher = more stigma. tgaywomen used as fallback for 2016",
    "0 = support marriage equality, 1 = oppose, NA = no opinion",
    "0 = support legal recognition, 1 = oppose, NA = no opinion",
    "0 = support adoption rights, 1 = oppose, NA = no opinion",
    "1 = support right to refuse service (high stigma), 0 = oppose",
    "1 = oppose trans bathroom access (high stigma), 0 = support"
  ),
  stringsAsFactors = FALSE
)

datatable(
  items_dict,
  rownames = FALSE,
  colnames = c("Variable", "Label", "Type", "Source Variable", "Scoring"),
  options = list(
    pageLength = 9,
    dom = "ft",
    columnDefs = list(
      list(width = "160px", targets = 0),
      list(width = "200px", targets = 1),
      list(width = "140px", targets = 2),
      list(width = "170px", targets = 3)
    )
  )
)
```

::: callout-important
`iat_sex_explicit_bel` (belief that sexuality is environmentally determined, from `sexualityorigin`) was planned but is **not included** — this variable does not appear in any Project Implicit sexuality dataset from 2016 onward.
:::

#### 3. State Filtering

After aggregation, state strings are normalized (uppercase, whitespace stripped, SPSS labels removed via `haven::zap_labels()`). Only the 50 U.S. states + D.C. are retained.

``` r
mutate(state = toupper(trimws(as.character(haven::zap_labels(state))))) |>
  filter(state %in% us_states_dc)
```

## Aggregation to State Level

### Unit of Analysis in Raw Data

Individual respondent (one row per person per year).

### Aggregation Method

| Feature | Decision | Rationale |
|---|---|---|
| Statistic | Mean | Standard for IAT D-score and Likert items aggregated to state level |
| Minimum n | None applied at this stage | Downstream analyses apply their own thresholds |
| Weighting | Unweighted | Project Implicit does not provide sampling weights |
| Missing states | Excluded | States with no respondents in a given year produce no row |

For each item, both the **mean** and a **respondent count** (`iat_sex_n_{item}`) are computed per state × year. The count columns are retained in the items dataset for downstream quality checks.

``` r
group_by(year, state) |>
  summarise(
    across(c(iat_sex_imp_d, ...), ~ sum(!is.na(.x)), .names = "iat_sex_n_{sub('iat_sex_', '', .col)}"),
    across(c(iat_sex_imp_d, ...), ~ mean(.x, na.rm = TRUE), .names = "{.col}"),
    .groups = "drop"
  )
```

## Output Dataset

### Variables in Final Output

#### Item-Level
**`iat_sexuality_items.Rds`** — item-level dataset (state × year)

| Variable | Type | Description | Scoring |
|---|---|---|---|
| `state` | chr | Two-letter state abbreviation | — |
| `year` | int | Calendar year | 2016–2025 |
| `iat_sex_imp_d` | dbl | Implicit D-score | Higher = more pro-straight bias |
| `iat_sex_exp_att` | dbl | Explicit attitude (7-pt) | Higher = more stigma |
| `iat_sex_exp_therm_gm` | dbl | Thermometer: gay men (reversed) | Higher = more stigma |
| `iat_sex_exp_therm_gw` | dbl | Thermometer: lesbian women (reversed) | Higher = more stigma |
| `iat_sex_exp_pol_marr` | dbl | Policy: marriage equality | 0–1; higher = more stigma |
| `iat_sex_exp_pol_legal` | dbl | Policy: legal recognition | 0–1; higher = more stigma |
| `iat_sex_exp_pol_adopt` | dbl | Policy: adoption rights | 0–1; higher = more stigma |
| `iat_sex_exp_pol_serv` | dbl | Policy: right to refuse service | 0–1; higher = more stigma |
| `iat_sex_exp_pol_trans` | dbl | Policy: trans bathroom access | 0–1; higher = more stigma |
| `iat_sex_n_{item}` | int | Respondent count per item | — |

#### Composite-level
**`iat_sexuality_indices.Rds`** — composite-level dataset (state × year)

| Variable | Type | Description | Components |
|---|---|---|---|
| `state` | chr | Two-letter state abbreviation | — |
| `year` | int | Calendar year | 2016–2025 |
| `iat_sex_implicit` | dbl | Implicit stigma index | `iat_sex_imp_d` |
| `iat_sex_explicit_therm` | dbl | Thermometer composite | Mean of `therm_gm`, `therm_gw` |
| `iat_sex_explicit_pol` | dbl | Policy composite | Mean of 5 policy items |
| `iat_sex_explicit` | dbl | Omnibus explicit index | Mean of all 7 explicit items |

Composites use `rowMeans(..., na.rm = TRUE)` so that state-years with partial item coverage still receive a score.

### Output Files

``` r
saveRDS(iat_sexuality_items,   here("data/clean_data/iat_sexuality_items.Rds"))
saveRDS(iat_sexuality_indices, here("data/clean_data/iat_sexuality_indices.Rds"))
```

**File locations:**

- `data/clean_data/iat_sexuality_items.Rds`
- `data/clean_data/iat_sexuality_indices.Rds`

## Known Limitations

- **Non-probability sample:** Project Implicit data come from self-selected volunteers. State-level estimates reflect the online population visiting the site, not representative state samples.
- **Implicit measure scope:** The D-score measures implicit associations for gay men only. Implicit attitudes toward lesbian women, bisexual, or other sexually diverse groups are not captured.
- **`iat_sex_explicit_bel` not available:** The planned belief-about-origin item (`sexualityorigin`) does not appear in any 2016+ dataset and is excluded.
- **2015 excluded:** Variable structure is incompatible with 2016+; a separate pipeline would be needed to incorporate it.
- **Thermometer variable rename:** `tgaywomen` was renamed `tgayleswomen` starting in 2017. The pipeline handles this via `coalesce()` but researchers should verify continuity if comparing pre- vs. post-2017 estimates.

## References

